humanCdGF_predict_cfu

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
remove old heatplot code (eval=F)

seed <- 1
n_trees <- 1001
cutoff <- 30

@@

```{r functions, eval = F}

heat.plot <- function(input_maxtrix, high_color = '#660000', low_color = '#FFFFFF', legend_title = NA) {
  long.df <- data.frame(input_maxtrix)
  colnames(long.df) <- c(1:10)
  long.df$community_day <- as.character(c(0:10))
  long.df %>% gather(cfu_day, variable, -community_day) %>% 
    ggplot(aes(cfu_day, community_day)) + 
    geom_tile(aes(fill = variable), color = 'white') +
    scale_fill_gradient(low = low_color, high = high_color, name = legend_title, limits = c(0,1)) +
    labs(x = 'Day (CFU)', y = 'Day (Community)') + 
    theme(axis.line.x = element_blank(), axis.line.y = element_blank(), axis.ticks = element_blank()) + 
    scale_x_discrete(expand = c(0, 0), limits = c('1','2','3','4','5','6','7','8','9','10')) + 
    scale_y_discrete(expand = c(0, 0), limits = rev(c('0','1','2','3','4','5','6','7','8','9','10'))) + 
    theme(legend.position = 'top', legend.key.width = unit(2, 'cm'))
}

n_features <- 10

rsq <- matrix(NA, nrow = 11, ncol = 10)
mse <- matrix(NA, nrow = 11, ncol = 10)
otu_votes <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes) <- colnames(rel_abund_shared_01)
importance.df <- c()
for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
    for(n in 1:10){
      cfu_shared.df <- meta_file %>% 
        filter(day == n) %>% 
        select(mouse_id, log_cfu) %>% 
        right_join(community.df) %>% 
        select(-mouse_id) %>% 
        filter(!is.na(log_cfu))

      set.seed(seed)
      rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
      importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
      top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
      set.seed(seed)
      rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                 names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
      rsq[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
      mse[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
      importance.df <- rbind(importance.df, c(RSQ = round(tail(rf_n_features$rsq, 1), 2), 
                      MSE = round(tail(rf_n_features$mse, 1), 2),
                      importance(rf_all)[,1]/max(importance(rf_all)[,1])))
    }
  }

importance.df <- data.frame(importance.df)
import_features <- importance.df[importance.df$RSQ >= 0.6 & importance.df$MSE <= 0.8, ]
import_features <- import_features[ , !colnames(import_features) %in% c('RSQ','MSE')]
med_import_feat <- apply(import_features, 2, median)
otus_above_20 <- import_features[,names(med_import_feat)[med_import_feat >= 0.2]]
top_n_overall <- names(otu_votes)[otu_votes >= cutoff]

rsq_n <- matrix(NA, nrow = 11, ncol = 10)
mse_n <- matrix(NA, nrow = 11, ncol = 10)
otu_votes_n <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes_n) <- colnames(rel_abund_shared_01)

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
  for(n in 1:10){
    cfu_shared.df <- meta_file %>% 
      filter(day == n) %>% 
      select(mouse_id, log_cfu) %>% 
      right_join(community.df) %>% 
      select(log_cfu, one_of(top_n_overall)) %>% 
      filter(!is.na(log_cfu))

    set.seed(seed)
    rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
    importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
    set.seed(seed)
    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                       names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
    rsq_n[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
    mse_n[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
    otu_votes_n <- otu_votes_n + 1*(colnames(rel_abund_shared_01) %in% top_important_OTU)
  }
}
taxa <- get_tax(1, names(otu_votes_n[otu_votes_n >= 1]), tax_file)
taxa_names <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
```

### OTUs selected by >`r cutoff`% of models generated by all OTU models 
###### Using the following bacteria: - r taxa_names

```{r top otus, fig.width=6, fig.height=6, eval = F}
heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2')))
#plot(otu_votes_n[otu_votes_n >= 1], xlab = 'OTU', ylab = 'Freq')
```

### CFU and Rel abundance plots

```{r fig.height=5, fig.width=8, eval = F}
meta_data %>% 
  filter(!is.na(day), !cage_id == 'NP1', 
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
    theme(legend.justification=c(1,0), legend.position=c(1,0), 
          legend.title=element_blank())
```

How does removing the low samples (Day 8 and CFUs around 0) affect prediction?

```{r identify reason for lowered R2, fig.height=5, fig.width=8, eval = F}
log_cutoff <- 2.5

meta_data %>% 
  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
  filter(!is.na(day), !cage_id == 'NP1', 
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
    theme(legend.justification=c(1,0), legend.position=c(1,0), 
          legend.title=element_blank()) + scale_y_continuous(limits = c(0,9))

meta_reduced <- meta_data %>% 
  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))

rsq_r <- matrix(NA, nrow = 11, ncol = 10)
otu_votes_r <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes_r) <- colnames(rel_abund_shared_01)

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
  for(n in 1:10){
    cfu_shared.df <- meta_reduced %>% 
      filter(day == n) %>% 
      select(mouse_id, log_cfu) %>% 
      right_join(community.df) %>% 
      select(log_cfu, one_of(top_n_overall)) %>% 
      filter(!is.na(log_cfu))

    set.seed(seed)
    rf_top <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
    importance_sorted <- sort(importance(rf_top)[,1], decreasing = T)
    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
    set.seed(seed)
    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                       names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
    importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
    top_OTU_final <- rownames(data.frame(head(importance_final, n_features)))
    rsq_r[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
    otu_votes_r <- otu_votes_r + 1*(colnames(rel_abund_shared_01) %in% top_OTU_final)
  }
}
taxa_r <- get_tax(1, names(otu_votes_r[otu_votes_r >= 1]), tax_file)
taxa_names_r <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
```

```{r fig.width=6,fig.height=6, eval = F}
heat.plot(rsq_r, '#003300', legend_title = expression(paste('R'^'2')))
```

Removing these data points improves predictability for day 6 and 8 but not the earlier days. If I increase the cutoff to remove all CFU below 5, there is a slight increase in day 3 cfu classification at the cost of decrease day 1 cfu classification.

### Is there any abnormalities in the OTUs of any these days?

```{r plot features,fig.width=8, fig.height=20, eval = F}
meta_data %>% 
  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
  filter(!cage_id == 'NP1', Early_Euth == FALSE, cdiff_strain == 431) %>% 
  full_join(rel_abund_shared_01) %>% 
  select(day, sample_id, mouse_id, cage_id, 
         one_of(names(otu_votes_r[otu_votes_r >= 1]))) %>% 
  gather(otu, rel_abund, Otu00002:Otu00250) %>% 
  filter(!is.na(day)) %>% 
  ggplot(aes(x=factor(day),y=rel_abund, color = cage_id)) +
    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
    facet_grid(otu ~., scales='free_y') +
    theme_bw() + labs(x = 'Day', y = 'Relative Abundance (%)')
```

### Does taking the median CFU improve predicatbility?

```{r predict median CFU heatmap, fig.width=6,fig.height=6, eval = F}

rsq_med <- matrix(NA, nrow = 11, ncol = 10)
otu_votes_med <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes_med) <- colnames(rel_abund_shared_01)
for(n in 1:10){
med_log_cfu <- meta_data %>% 
  filter(day %in% c(n:10), !cage_id == 'NP1', 
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  select(mouse_id, cage_id, log_cfu) %>% 
  group_by(mouse_id, cage_id) %>% 
  summarise(log_cfu_med = median(log_cfu)) %>% ungroup()
for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
  cfu_shared.df <- med_log_cfu %>% 
      select(mouse_id, log_cfu_med) %>% 
      right_join(community.df) %>% 
      select(-mouse_id) %>% 
#      select(log_cfu_med, one_of(top_n_overall)) %>% 
      filter(!is.na(log_cfu_med))
    set.seed(seed)
    rf_top <- randomForest(log_cfu_med ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
    importance_sorted <- sort(importance(rf_top)[,1], decreasing = T)
    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
    set.seed(seed)
    rf_n_features <- randomForest(log_cfu_med ~ ., data=cfu_shared.df[ ,c( 'log_cfu_med',
                                                                       names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
    importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
    top_OTU_final <- rownames(data.frame(head(importance_final, n_features)))
    rsq_med[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
    otu_votes_med <- otu_votes_med + 1*(colnames(rel_abund_shared_01) %in% top_OTU_final)
}
}

heat.plot(rsq_med, '#003300', legend_title = expression(paste('R'^'2'))) + 
  ggtitle('Day CFU is median value of Day n to Day 10')

###### Re-runnning the program and only focusing on the OTUs predicting day 9/10 to see if there is consistence in the predictive features gives the following selection  (including all samples):  
Otu00004 Otu00005 Otu00015 Otu00019 Otu00030 Otu00199 Otu00200 Otu00250   
      14       17       13       15       13       16       17       19

```{r boruta feature selection, eval = F}
bor_all <- c()

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
    for(n in 9:10){
      cfu_shared.df <- meta_file %>% 
        filter(day == n) %>% 
        select(mouse_id, log_cfu) %>% 
        right_join(community.df) %>% 
        select(-mouse_id) %>% 
        filter(!is.na(log_cfu))

      set.seed(seed)
      bor_all[[(m*2)+(n-8)]] <- Boruta(log_cfu ~ ., data = cfu_shared.df, maxRuns = 2000)
    }
  }

boruta.votes <- rep(0, ncol(rel_abund_shared_01))
names(boruta.votes) <- colnames(rel_abund_shared_01)
for(i in 1:22){
  boruta.votes <- boruta.votes + 1*(bor_all[[i]]$finalDecision   == 'Confirmed')
}
plot(boruta.votes)
boruta.votes[boruta.votes >= 16]

colnames(otus_above_20)
```

Boruta Confirmed the following OTUs as important for predicting day 9/10 cfu:   
Otu00004 Otu00005 Otu00015 Otu00030 Otu00081 Otu00199 Otu00200 Otu00250 Otu00297  
      21       22       21       21       21       22       22       22       20

Selecting OTUs through collecting the features from the most predictive community/cfu models (R^2 >= 0.6 and MSE <= 0.8), then converting all % Increase in MSE to relative values and taking the median value of of each OTU, then selecting OTUs that fall above the median value results in the following OTUs:  
"Otu00001" "Otu00002" "Otu00004" "Otu00005" "Otu00012" "Otu00014" "Otu00015" "Otu00016" "Otu00019" "Otu00028" "Otu00030" "Otu00048" "Otu00081" "Otu00101" "Otu00118" "Otu00199" "Otu00200" "Otu00250" "Otu00297"
```{r }

>>>>

seed <- 1
n_trees <- 1001

@@
